---
title: "Additional report stuff"
author: "Jonathan Marshall"
date: "18 January 2018"
output: html_document
---

```{r setup, include=FALSE}
library(dplyr)
library(tidyr)
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
prev2009 <- read.csv("../data/2009_sample_size_calcs.csv")
dat2017 <- read.csv("../data/2017_nmd_prev.csv") %>%
  mutate(Prevalence = round(Positives/Samples * 100, 1)) %>%
  select(Animal, Genus, Date, Samples, Positives, Prevalence)
```

## Samples required for isolates

As noted earlier, the prevalences from the 2009 study represent true prevalences only for *Enterococcus* across all animal groups, and *Campylobacter* for very young calves and pigs. The *E.coli* results for all animal groups, and *Campylobacter* results for poultry are both conditional on an NMD positive sample being sent for additional isolation.

To assess true prevalence, and thus estimate sample sizes, additional data, including *Salmonella*, was obtained from the NMD (Gail Duncan, pers comm) for the 3rd quarter of 2017 and is presented below. Note the low prevalence in general of Salmonella, and the lower prevalence of E. coli among adult cattle.

```{r}
dat2017 %>% kable
```

Combining these NMD results with the prevalence data (which represents the likely success of obtaining an isolate given an NMD positive) from the 2009 study yields the following estimates of prevalence.

```{r}
prev2009 %>%
  select(Animal, Genus, CI) %>%
  spread(Genus, CI) %>%
  kable
```

The corresponding sample size required to produce 300 isolates for each genus is given below, computed using the 25th percentile of estimated prevalence to ensure we're 75% likely to get the required number of isolates. As can be seen, the sample sizes required for Campylobacter on non-poultry sources are infeasible.

```{r}
prev2009 %>%
  select(Animal, Genus, Samples) %>%
  spread(Genus, Samples) %>%
  kable
```

## Sampling frequency

If isolates are exclusively sourced from NMD, then a consideration is that samples in NMD are at fixed intervals and do not vary based on throughput. To achieve a sampling plan where samples are in proportion to throughput, the 2009 survey requested samples from LAS labs for different plants at different frequencies, ranging from weekly through to monthly.

A similar scheme would be required for a further sampling round. The number of samples to collect excluding non-poultry *Campylobacter* are within the numbers collected by NMD for routine testing, and hence LAS labs might again be utilised to gather the required rinsates, plates, or petrifilms for further isolation and antimicrobial susceptibility testing.